---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CODECtools

<!-- badges: start -->
[![R-CMD-check](https://github.com/geomarker-io/CODECtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/geomarker-io/CODECtools/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of CODECtools is to support CODEC data infrastructure through:

- capturing, creating, and accessing metadata for tabular data in R
- importing and exporting tabular data resources (CSV file) with metadata (YAML file)
- defining CODEC data format specifications (as vignette in package)
- providing tools and workflows to validate CODEC data and contribute it to CODECdata repository
- creation of online data catalog

## Installation

You can install the development version of CODECtools from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("geomarker-io/CODECtools")
```

## Metadata Example

Here is a basic example of adding metadata to data in R, saving it to disk, and reading it back into R with the metadata:

```{r}
library(CODECtools)

my_mtcars <-
  mtcars |>
    add_attrs(name = "Motor Trend Cars", year = "1974") |>
    add_type_attrs() |>
    add_col_attrs(mpg, name = "MPG", description = "Miles Per Gallon")

get_descriptors(my_mtcars) |>
  knitr::kable()

get_schema(my_mtcars) |>
  knitr::kable()

save_tabular_data_resource(my_mtcars)

## yaml::yaml.load_file("tabular-data-resource.yaml")

```

