---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# CODECtools

<!-- badges: start -->
[![R-CMD-check](https://github.com/geomarker-io/CODECtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/geomarker-io/CODECtools/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of CODECtools is to support CODEC data infrastructure through:

- capturing, creating, and accessing metadata for tabular data in R
- importing and exporting tabular data resources (CSV file) with metadata (YAML file)
- defining CODEC data format specifications (as vignette in package)
- providing tools and workflows to validate CODEC data and contribute it to CODECdata repository
- creation of online data catalog

## Installation

You can install the development version of CODECtools from [GitHub](https://github.com/) with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("geomarker-io/CODECtools")
```

## A Simple Metadata Example

```{r}
library(CODECtools)
```

We will create a simple dataset here for this example:

```{r}
d <-
  tibble::tibble(
    id = c("A01", "A02", "A03"),
    date = as.Date(c("2022-07-25", "2018-07-10", "2013-08-15")),
    measure = c(12.8, 13.9, 15.6),
    rating = factor(c("good", "best", "best"), levels = c("good", "better", "best")),
    ranking = as.integer(c(14, 17, 19)),
    impt = c(FALSE, TRUE, TRUE)
  )
```

When creating a tabular dataset in R, data-specific metadata (i.e., "descriptors") can be stored in the attributes of the R object (e.g., a data.frame or tibble).

```{r}
d <- d |>
  add_attrs(
    name = "mydata",
    title = "My Data",
    license = "MIT",
    url = "https://geomarker.io/CODECtools"
  )
```

Note that this doesn't change any of the data values. In R, an object's attributes are stored with it as a list. Some attributes (`?attributes`) are treated specially by R (e.g., `class`, `names`, `row.names`, `comment`) and usually shouldn't be modified.  Although *all* attributes (including the ones we added above) are available as a list (`?attributes`), we can use a function to extract only the attributes that represent metadata descriptors as a tibble.

```{r}
get_descriptors(d) |>
  knitr::kable()
```

Similarly, we can add column-specific attributes (i.e., "schema"). These metadata functions follow the tidy design principles, making it simple to expressively and concisely add metadata using pipes:

```{r}
d <-
  d |>
  add_col_attrs(id, title = "Identifier", description = "unique identifier") |>
  add_col_attrs(date, title = "Date", description = "date of observation") |>
  add_col_attrs(measure, title = "Measure", description = "measured quantity") |>
  add_col_attrs(rating, title = "Rating", description = "ordered ranking of observation") |>
  add_col_attrs(ranking, title = "Ranking", description = "rank of the observation") |>
  add_col_attrs(impt, title = "Important", description = "true if this observation is important")
```

Automatically add `name`, `type` and `enum` schema to each column in the data based on their class:

```{r}
d <- add_type_attrs(d)
```

Like for descriptors, there is a helper function to retrieve schema as a tibble.  Specify `bind = TRUE` to combine the list of column-specific attribute data frames for each column into one wider data frame:

```{r}
get_schema(d, bind = TRUE) |>
  knitr::kable()
```

Once metadata is set in the tibble's attributes, we can save the tabular data resource as a CSV file with an accompanying tabular-data-resource.yaml:

```{r}
write_tdr_csv(d)
``` 

The `name` attribute of the supplied tibble is used as the name of a newly created folder *and* CSV file containing the data. Metadata extracted from the supplied tibble's attributes is saved in a `tabular-data-resource.yaml` file that lives alongside the data file in the newly created directory:

```{r}
fs::dir_tree("mydata")

```

We can then read this tabular-data-package back into R and restore its attributes, as well as its column classes:

```{r}
mydata <- read_tdr_csv("mydata")
mydata
```

```{r, include = FALSE}
fs::dir_delete("mydata")
```
