---
title: "Working with a Tabular Data Resource in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with a Tabular Data Resource in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(CODECtools)
```

Inside R, metadata lives in the attributes of the data.frame and its columns. We can add and change these with several helper functions used in the example below: `add_attrs()`, `add_col_attrs()`, `add_type_attrs()`. Using these functions to set attributes in R means we can do so reproducibly and changes to the metadata are tracked alongside the R script that creates the data. This prevents a disconnect between data and metadata, but also allows for computing on the metadata to use it to create richer documentation. ^[Alternatively, we can edit the metadata yaml file directly and track changes there.  Reading the data CSV file back in to R using a function that also reads metadata (`read_tdr_csv()`) will update the attributes of the data.frame based on changes in the metadata file.]

## Creating Data

We will create a simple dataset here for this example:

```{r}
d <-
  tibble::tibble(
    id = c("A01", "A02", "A03"),
    date = as.Date(c("2022-07-25", "2018-07-10", "2013-08-15")),
    measure = c(12.8, 13.9, 15.6),
    rating = factor(c("good", "best", "best"), levels = c("good", "better", "best")),
    ranking = as.integer(c(14, 17, 19)),
    impt = c(FALSE, TRUE, TRUE)
  )
```

## Adding Metadata

When creating a tabular dataset in R, data-specific metadata (i.e., "descriptors") can be stored in the attributes of the R object (e.g., a data.frame or tibble).

```{r}
d <- d |>
  add_attrs(
    name = "mydata",
    title = "My Data",
    license = "MIT",
    url = "https://geomarker.io/CODECtools"
  )
```

Note that this doesn't change any of the data values. In R, an object's attributes are stored with it as a list. Some attributes (`?attributes`) are treated specially by R (e.g., `class`, `names`, `row.names`, `comment`) and usually shouldn't be modified.  Although *all* attributes (including the ones we added above) are available as a list (`?attributes`), we can use a function to extract only the attributes that represent metadata descriptors as a tibble.

```{r}
glimpse_attr(d) |>
 knitr::kable()
```

Similarly, we can add column-specific attributes (i.e., "schema"). These metadata functions follow the tidy design principles, making it simple to expressively and concisely add metadata using pipes:

```{r}
d <-
  d |>
  add_col_attrs(id, title = "Identifier", description = "unique identifier") |>
  add_col_attrs(date, title = "Date", description = "date of observation") |>
  add_col_attrs(measure, title = "Measure", description = "measured quantity") |>
  add_col_attrs(rating, title = "Rating", description = "ordered ranking of observation") |>
  add_col_attrs(ranking, title = "Ranking", description = "rank of the observation") |>
  add_col_attrs(impt, title = "Important", description = "true if this observation is important")
```

Automatically add `name`, `type` and `enum` schema to each column in the data based on their class:

```{r}
d <- add_type_attrs(d)
```

Like for descriptors, there is a helper function to retrieve schema as a tibble: 

```{r}
glimpse_schema(d) |>
  knitr::kable()
```

## Writing a tabular-data-resource to disk

Once metadata is set in the tibble's attributes, we can save the tabular data resource as a CSV file with an accompanying tabular-data-resource.yaml:

```{r}
write_tdr_csv(d)
``` 

The `name` attribute of the supplied tibble is used as the name of a newly created folder *and* CSV file containing the data. Metadata extracted from the supplied tibble's attributes is saved in a `tabular-data-resource.yaml` file that lives alongside the data file in the newly created directory:

```{r}
fs::dir_tree("mydata")
```

## Reading a tabular-data-resource from disk

We can then read this tabular-data-package back into R and restore its attributes, as well as its column classes:

```{r}
mydata <- read_tdr_csv("mydata")
mydata
```

```{r, include = FALSE}
fs::dir_delete("mydata")
```
